### 2022-05-19 paper.Rmd
writing

    Anticlustering refers to a collection of methods that partition data sets into disjoint groups while aiming for high between-group similarity and high within-group heterogeneity. Anticlustering is usually approached by maximzing instead of minimizing a clustering objective such as the k-means criterion. Introducing k-plus, this paper presents an extension of k-means to maximize between-group similarity. While k-means anticlustering minimizes between-group-differences only with regard to mean attribute values, k-plus anticlustering can be used to minimize between-group-differences with regard to the variance, covariance structure, and higher order moments. To minimize differences in variance, each variable is duplicated by computing the squared deviation between the original values and the variable's mean. Because the variance is defined as the expected value of the squared deviations from the mean, appending these duplicated variables to the data set---and then applying standard k-means anticlustering---suffices to equalize the variance between groups. Following a similar logic, differences in skewness, kurtosis and covariances can be minimized. Thus, while k-plus is a separate criterion for anticlustering, it can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples on real data show that k-plus anticlustering achieves high similarity with regard to mean attribute values as well as the variance and higher order moments, usually with no noticable sacrifice to either criterion. The k-plus extension is therefore preferred over classical k-means anticlustering for optimizing between-group similarity. Example are given on how k-plus anticlustering can be applied using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering methods can be distinguished by the input data and the objective function that is used to optimize group between-group similarity or within-group heterogeneity. Some anticlustering objectives quantify group similarity on the basis of pairwise dissimilarity rating such as the Euclidean distance. The most prominent criterion on the basis of pairwise dissimilarity ratings is the *diversity*, which is the total sum of dissimilarities between any elements within the same group. Maximizing the diversity optimizes within-group heterogeneity, which simultaneously leads to similar groups---if all groups have the same size. Anticlustering objective functions usually represent the reversal of a clustering objective, only the direction of the optimization is changed. Because the diversity is the reversal of the clustering editing objective [@zahn1964; @shamir2004vf], @papenberg2020 used the term *anticluster editing* to refer to anticlustering approaches that optimize the diversity. Another important criterion based on pairwise dissimilarities is the *dispersion*, i.e., the minimum dissimilarity between any two elements within the same group [@fernandez2013; @brusco2019]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible.

Oftentimes, researchers work with attribute values [or *features*; @dry2009] instead of dissimilarity ratings. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing].  For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to partition the elements into groups. A different approach to anticlustering directly works with the attribute data by maximizing the k-means criterion. K-means is probably the best-known clustering method. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful anticlustering tool by Späth [-@spath1986] and Valev [-@valev1998; -@valev1983]. In k-means clustering, the sum of the squared Euclidean distances between elements and the center of the cluster they are assigned to is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. For the anticlustering application, the k-means criterion is maximized instead, thereby achieving similarity between groups. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. Hence, groups are similar with regard to the mean of the distribution of the numeric attributes wheras other parameters---such as the variance---are ignored. When requiring that groups are overall similar to each other, this focus may be misguided as similar means can be obtained even when the underlying distributions are different [e.g., @anscombe1973]. @papenberg2020, showed that maximizing the diversity (based on the Euclidean distance) is better suited to minimize difference with regard to both the mean and the variance of the data; even an entirely random split usually resulted in more similar variances because k-means solely focuses on means.

To optimize overall group similarity---and hence to overcome the limitations of k-means anticlustering---I present k-plus anticlustering in this paper. The k-plus criterion is an extension of the k-means objective that includes a term that quantifies similarity with regard to the variance, covariances, and higher order moments of data across groups. Interestingly, this extended criterion can be understood as an augmentation of the original data matrix, leaving the original k-means objective unchanged. I show that the k-plus criterion outperforms other anticlustering criteria when the goal is to minimize differences with regard to the mean and the variance of groups. In most cases, both objectives can be optimized satisfactorily at the same time without any loss on either. Moreover, in two pratical examples I illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

## The k-plus family of anticlustering objectives

Some clarifications with respect to nomenclature are needed. The k-plus criterion is not a single objective function for anticlustering, but instead a familiy of objectives that extend the standard k-means objective ($SSE$) to minimize differences between groups with regard to several characterizations of a data set. That is: not just the mean, but also the variance, covariances, or higher order moments such as skewness and kurtosis. Which of these objectives are pursued depends on a researchers needs. In the most basic---and arguably most important---case, k-plus anticlustering can be used to not just minimize differences with regard to the mean of variables but also their variance. Measures of location (mean) and spread (variance / standard deviation) are arguably most important to researchers when they "eyeball" their data and they are routinely reported as characterizations of a data set. Hence, by default, the k-plus anticlustering objective implies that means and variances should be equal between groups.

While less routinely reported, skewness are kurtosis are also important descriptive characterizations of a data set [@decarlo1997; @westfall2014]. To optimize overall similarity between groups, it may thus be desirable to also minimize differences with regard to these distribution moments. In principle, any higher order moments can also be included as part of the k-plus objective. Note, however, that the inclusion each additional moment requires appending as many features to the data set as there were to begin with.

K-plus anticlustering can also be used to obtain similar covariance structures among groups. This approach may be of interest in cross validation applications where prediction accuracy depends on the correlations among features and criteria [@zeng2000; @papenberg2020]. Note that each additional objective that is pursued with k-plus anticlustering comes with a cost. It becomes more difficult to satisfactorily fulfil many criteria at the same time. As the simulation study below will show, including many k-plus objectives at the same time becomes more appropriate when the data set becomes larger. In the end, it is a tradeoff between fulfilling few objectives to a precise degree or fulfilling many objectives to a less compelling degree.

@@

To systematically evaluate the new k-plus criterion, I compared three anticlustering approaches in a simulation study: (a) k-means anticlustering, optimizing the $\mathit{SSE}$; (b) k-plus anticlustering, optimizing the $SSE_{kplus}$; (c) anticluster editing, optimizing the diversity criterion. All methods used a $z$-standardization of the data sets as a preprocessing step. In the case of k-plus anticlustering, this means that the augmented input data (i.e., the data table that contains the original data and the squared distance data) were subjected to a standardization, which is recommended.

The simulation used the statistical programming environment R [@R-base]. The R package `anticlust` [Version 0.5.6; @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R packages `dplyr` [Version 1.0.7; @R-dplyr], `ggplot2` [Version 3.3.5; @R-ggplot2], `papaja` [Version 0.1.0.9997; @R-papaja] and `tidyr` [Version 1.1.3; @R-tidyr] were used for data processing and result presentation. The simulation study is fully reproducible via code and data that is accessible from the accompanying OSF repository (**link TODO**).

@@

First, I generated 10,000 data sets which were subsequently processed by to the three competing anticlustering methods. The following properties were determined randomly for each data set: (a) the sample size $N$ varied between 24 and 300 with intermediate steps of 12 so that each data set could be split evenly into $K = 2$, $3$, and $4$ groups; (b) the number of features varied between 1 and 10; (c) the distribution of features was randomly determined to be either (i) uniform in [0, 1], (ii) a standard normal distribution with *M* = 0 and *SD* = 1, or (iii) a wider normal distribution with *M* = 0 and *SD* = 2. For a given data set, all $M$ features were independently sampled from the given distribution and no correlations were assumed between features. All three anticlustering methods were applied to each of the 10,000 data sets using $K = 2$, $K = 3$ and $K = 4$. Groups sizes were always equal.

To optimize the three anticlustering criteria, a local-maximum search was applied that systematically swaps elements between groups to gradually improve the objective function [Method LCW in @weitz1998]. Building on an initial random assignment of elements to groups, elements are swapped between groups in such a way that each swap improves the objective criterion by the largest amount that is possible. That is, for each element, all possible swaps are simulated and the resulting objective values are stored. After each exchange has been tested, the one exchange is realized that improves the objective the most. No swap is realized if no swap improves the criterion. The exchange procedure is repeated for each element. Once the last element has been processed, the procedure restarts at the beginning and once again iterates through all elements. The search stops as soon as an iteration through the entire set no longer yields any improvement, i.e., as soon as a local maximum has been found. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. The simulation employed five repetitions of the local maximum search.

@@

Two primary performance criteria were assessed. First, for k-plus anticlustering, I checked how well the identified solution approximated the ordinary k-means criterion. Specifically, I computed the $\mathit{SSE}$ for the solution that was returned by k-means anticlustering---that directly optimizes $\mathit{SSE}$---and for the solution that was returned by k-plus anticlustering that only indirectly optimizes $\mathit{SSE}$ through $SSE_{kplus}$. If optimizing $SSE_{kplus}$ simultaneously optimizes the standard $\mathit{SSE}$ to a satisfactory degree, we may conclude that simultaneously striving for similar means and variances is feasible by optimizing the k-plus criterion.

Second, to compare the three anticlustering criteria with regard to their ability to create similar groups, I computed the means and standard deviations and assessed the degree to which discrepancies occurred between groups. To this end, I then computed the difference between the minimum and maximum value of the means and standard deviations between groups, for each of the $M$ features. The mean of the differences across all features quantified the total dissimilarity in means ($\Delta M$) and standard deviations ($\Delta \mathit{SD}$). Therefore, lower values of $\Delta M$ and $\Delta \mathit{SD}$ indicated that groups were more similar with regard means or standard deviations, respectively.

>>>>

    Anticlustering refers to a collection of methods that partition data sets into disjoint groups while aiming for high between-group similarity and high within-group heterogeneity. Anticlustering is usually approached by maximzing instead of minimizing a clustering objective such as the k-means criterion. Introducing k-plus, this paper presents an extension of k-means to maximize between-group similarity. While k-means anticlustering minimizes between-group-differences only with regard to the mean of attribute values, k-plus anticlustering can be used to minimize between-group-differences with regard to the variance, higher order moments, and covariance structure. To minimize differences in variance, each variable is duplicated by computing the squared deviation between the original values and the variable's mean. Because the variance is defined as the expected value of the squared deviations from the mean, appending these duplicated variables to the data set---and then applying standard k-means anticlustering---suffices to equalize the variance between groups. Following a similar logic, differences in skewness, kurtosis and covariances can be minimized. Thus, while k-plus is a separate criterion for anticlustering, it can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples on real data show that k-plus anticlustering achieves high similarity with regard to mean attribute values as well as the variance and higher order moments, usually with no noticable sacrifice to either criterion. The k-plus extension is therefore preferred over classical k-means anticlustering for optimizing between-group similarity. Example are given on how k-plus anticlustering can be applied using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

Anticlustering methods can be distinguished by the input data and the objective function that is used to optimize group between-group similarity / within-group heterogeneity. Some anticlustering objectives quantify group similarity on the basis of pairwise dissimilarity rating such as the Euclidean distance. The most prominent criterion on the basis of pairwise dissimilarity ratings is the *diversity*, which is the total sum of dissimilarities between any elements within the same group. Maximizing the diversity optimizes within-group heterogeneity, which simultaneously leads to similar groups---if all groups have the same size. Anticlustering objective functions usually represent the reversal of a clustering objective, only the direction of the optimization is changed.  Another important criterion based on pairwise dissimilarities is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013; @brusco2019]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible.

Oftentimes, researchers work with attribute values [or *features*; @dry2009] instead of dissimilarity ratings. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing].  For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to partition the elements into groups. A different approach to anticlustering directly works with the attribute data by maximizing the k-means criterion. K-means is probably the best-known clustering method. Reversing the k-means objective function---using maximization instead of minimization---has been independently recognized as a useful anticlustering tool by Späth [-@spath1986] and Valev [-@valev1998; -@valev1983].

In k-means clustering, the sum of the squared Euclidean distances between elements and the center of the cluster they are assigned to is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. For the anticlustering application, the k-means criterion is maximized instead, thereby achieving similarity between groups. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. Hence, groups are similar with regard to the mean of the distribution of the numeric attributes wheras other parameters---such as the variance---are ignored. When requiring that groups are overall similar to each other, this focus may be misguided as similar means can be obtained even when the underlying distributions are different [e.g., @anscombe1973]. @papenberg2020 showed that maximizing the diversity (using the Euclidean distance) is better suited to minimize difference with regard to both the mean and the variance of the data; even an entirely random split usually resulted in more similar variances because k-means solely focuses on the features' means.

To optimize overall group similarity---and hence to overcome the limitations of k-means anticlustering---this paper presents k-plus anticlustering. The k-plus criterion is an extension of the k-means objective that includes one or several terms to quantify similarity with regard to other important characteristics of a data set rather than only the mean. Hence, the k-plus criterion is not a single objective function for anticlustering, but instead a familiy of objectives that extend the standard k-means objective to minimize differences between groups with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. Which of these objectives is pursued depends on a researchers' needs. In the most basic---and arguably most important---case, k-plus anticlustering can be used to not just minimize differences with regard to the mean of variables but also their variance. Measures of location and spread are arguably most important to researchers when they "eyeball" their data and they are routinely reported as characterizations of a data set. Hence, by default, the k-plus objective is refered to when considering the mean and the variance of variables in the anticlustering process.

While less routinely reported, skewness are kurtosis are also important descriptive characterizations of a data set [@decarlo1997; @westfall2014]. To optimize overall similarity between groups, it may thus be desirable to also minimize differences with regard to these distribution moments. In principle, any other higher order moments can also be included as part of the k-plus objective. Futhermore, k-plus anticlustering can also be used to obtain similar covariance structures among groups. This approach may be of interest in cross validation applications where prediction accuracy depends on the correlations among features and criteria [@zeng2000; @papenberg2020]. Interestingly, in each case, the k-plus criterion can be reduced to an augmentation of the original data matrix, leaving the original k-means objective unchanged.

After mathematically introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups that have minimal differences according to multiple distribution characteristics. Fulfilling multiple criteria at the same time becomes more feasible as the size of the data set increases. In two pratical examples I illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

$$
\frac{1}{N} \sum\limits_{i=1}^{N} (x_i - \overline{\mathbf{x}}) (y_i -\overline{\mathbf{y}}) 
$$

@@

Three k-plus criteria were implemented in a simulation study: (i) standard *k-plus anticlustering*, minimizing differences with regard to the mean and variance; (ii) *k-plus skew/kurtosis*, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) *k-plus correlation*, minimizing differences with regard to the mean, variance and covariance structure. The k-plus methods were compared to (a) standard k-means anticlustering, (b) diversity anticlustering using the Euclidean distance, and (c) a simple random allocation of elements to groups. All anticlustering methods used a $z$-standardization of the data sets as a preprocessing step. In the case of k-plus anticlustering, this means that the augmented input data (i.e., the data table that contains the original data and the features that were added to incorporate the additional optimization criteria) were subjected to a standardization, which is recommended.

The simulation used the statistical programming environment R [@R-base]. The R package `anticlust` [Version 0.6.1; @R-anticlust; @papenberg2020] was used to optimize the anticlustering objectives. The R packages `dplyr` [Version 1.0.7; @R-dplyr], `ggplot2` [Version 3.3.5; @R-ggplot2], `papaja` [Version 0.1.0.9997; @R-papaja] and `tidyr` [Version 1.1.3; @R-tidyr] (**TODO: citation faux**) were used for data processing and result presentation. The simulation study is fully reproducible via code and data that is accessible from the accompanying OSF repository (**link TODO**).

The same local maximum search was applied to optimize each of the competing five anticlustering methods [Method LCW in @weitz1998]. Building on an initial random assignment of elements to groups, elements are swapped between groups in such a way that each swap improves the objective criterion by the largest amount that is possible. That is, for each element, all possible swaps are simulated and the resulting objective values are stored. After each exchange has been tested, the one exchange is realized that improves the objective the most. No swap is realized if no swap improves the criterion. The exchange procedure is repeated for each element. Once the last element has been processed, the procedure restarts at the beginning and once again iterates through all elements. The search stops as soon as an iteration through the entire set no longer yields any improvement, i.e., as soon as a local maximum has been found. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. The simulation employed five repetitions of the local maximum search.

@@

First, I generated 10,000 data sets following a normal distribution. The data sets were subsequently processed by the competing methods. The following properties were determined randomly for each data set: (a) the sample size $N$ varied between 24 and 300 with intermediate steps of 12 so that each data set could be split evenly into $K = 2$, $3$, and $4$ groups; (b) the number of features varied between 2 and 5; (c) the standard deviation of all features was set to 1, 2 or 3 (the mean was always zero); (d) the correlation between all features was r = 0, r = .1, r = .2, r = .3, r = .4 or r = .5. All five anticlustering methods were applied to each of the 10,000 data sets using $K = 2$, $K = 3$ and $K = 4$ (i.e., each method was applied 30,000 times). Groups sizes were always equal.

@@

To compare the five anticlustering criteria with regard to their ability to create similar groups, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between features across groups.

**todo hier weitermachen**

I then computed the difference between the minimum and maximum value of the means and standard deviations between groups, for each of the $M$ features. The mean of the differences across all features quantified the total dissimilarity in means ($\Delta M$) and standard deviations ($\Delta \mathit{SD}$). Therefore, lower values of $\Delta M$ and $\Delta \mathit{SD}$ indicated that groups were more similar with regard means or standard deviations, respectively.
